---
title: "Initial clustering for DISCOV-R"
author: "Virginia S. Muir"
date: "9/5/2018"
output: pdf_document
---

## ============================
## Prepare environment
## ============================

```{r set_up, echo=FALSE, include=FALSE}
# Set up the environment with required packages
library(tidyverse)
library(flowStats)
library(flowCore)
library(Rphenograph)
library(Rtsne)
library(igraph)
library(ggthemes)
library(gridExtra)

# Session options
options(stringsAsFactors = FALSE)
set.seed(12345)

# Set up the ggplot default params
theme_set(theme_bw(12) + 
          theme(panel.grid.major = element_blank(), 
                panel.grid.minor = element_blank(),
                plot.title = element_text(size=15, face="bold", margin = margin(10,0,10,0)),
                axis.text.x = element_text(angle=45, hjust = 1)))
cb_pal <- colorblind_pal()(8)[-1] 

# Generate file structure in which to save R data and exports
dir.create("Output")
dir.create(file.path("Output", format(Sys.Date(), "%y%m%d")))
dir.create(file.path("Output", format(Sys.Date(), "%y%m%d"), "R_objects"))
dir.create(file.path("Output", format(Sys.Date(), "%y%m%d"), "tSNE_plots"))

fname_prefix_R <- file.path("Output", format(Sys.Date(), "%y%m%d"), "R_objects", 
                               format(Sys.Date(), "%y%m%d"))
fname_prefix_tsne <- file.path("Output", format(Sys.Date(), "%y%m%d"), "tSNE_plots", 
                               format(Sys.Date(), "%y%m%d"))

```

## ============================
## Load & format data
## ============================

```{r import_and_format_data}
## There are a couple of ways you can import your data, depending on which data you'd like to pull in.
# List all files in the data folder named FCS
all_files <- list.files("FCS", full.names = T, recursive = T)

# List all files that end in .fcs anywhere in your working directory & its sub-directories
#all_files <- list.files(pattern = ".fcs", full.names = T, recursive = T)

# Subset to total CD8+ files (this looks in your list of files for perfect matches to the "CD8_T_Cells" string - capitilization must be consistent)
cd8_files <- all_files[str_detect(all_files, pattern = "CD8_T_Cells")]

# Subset to tmr+ files
islet_tmr_files <- all_files[str_detect(all_files, "PPI_GAD_IAPP_IGRP.fcs")]
insulin_tmr_files <- all_files[str_detect(all_files, "Ins.fcs")]
viral_tmr_files <- all_files[str_detect(all_files, "CMV_EBV.fcs")]

# Read in CD8+ and tmr+ fcs files to build flowSets
build_fcsList <- function(file_list) {
  tmp_list <- list()
  for (fcs in file_list) {
    if (str_detect(fcs, "xpb") == T) {
      fcs_name <- paste(str_match(fcs, "xpb[0-9]+_(.*?)_")[2])
    } else {
      fcs_name <- paste("T1D-", str_match(fcs, "A02_([0-9]{5})_0")[2], sep="")
    }
    tmp_list[[fcs_name]] <- read.FCS(fcs, transformation = F)
  }
  tmp_list
}

cd8_files <- build_fcsList(cd8_files)
islet_tmr_files <- build_fcsList(islet_tmr_files)
insulin_tmr_files <- build_fcsList(insulin_tmr_files)
viral_tmr_files <- build_fcsList(viral_tmr_files)
#fcsNames = names(cd8_files)

# Pull in key to correct inconsistent labels & to specify markers of interest
key = read.csv("MarkerKey.csv", header = T)

#Process data
process_data <- function(fcs){

  # Tidy marker names
  fcs_key <- pData(parameters(fcs))
  pData(parameters(fcs))$desc <- toupper(key$desc[match(fcs_key$name, key$name)])

  # This changes parameters(fcs)$name, featureNames(fcs), and colnames(fcs) - aka events colnames - all in one fell swoop.
  colnames(fcs) = make.names(pData(parameters(fcs))$desc)

  # Remove markers that aren't in marker key list of 24 (i.e. duplicated NAs)
  fcs = fcs[,!(duplicated(colnames(fcs)) | duplicated(colnames(fcs), fromLast = T))] 
  fcs = fcs[, order(colnames(fcs))]
}

cd8_files <- lapply(cd8_files, process_data)
islet_tmr_files <- lapply(islet_tmr_files, process_data)
insulin_tmr_files <- lapply(insulin_tmr_files, process_data)
viral_tmr_files <- lapply(viral_tmr_files, process_data)


```


```{r transform_and_extract_data}
# Merge total and tmr+ expression data
all_merged = cd8_files
for(sample in unique(names(cd8_files))){
  exprs(all_merged[[sample]]) = rbind(flowCore::exprs(cd8_files[[sample]]), 
                                    if(!is.null(islet_tmr_files)) flowCore::exprs(islet_tmr_files[[sample]]), 
                                    if(!is.null(insulin_tmr_files)) flowCore::exprs(insulin_tmr_files[[sample]]), 
                                    if(!is.null(viral_tmr_files)) flowCore::exprs(viral_tmr_files[[sample]]))
}

# Transform the data
asinh_tfm_data <- function(fcs){
  # Arcsinh transform all remaining columns; Use b = 1/5 for CyTOF data, b = 1/150 for Flow data
  tl <- transformList(colnames(fcs), arcsinhTransform(a=0, b=1/5), transformationId="asinh")
  fcs = transform(fcs, tl) 
}

all_tfm <- lapply(all_merged, asinh_tfm_data)

tfm_fs <- as(all_tfm, "flowSet")

# Extract expression data and label Tmr+ events
mergedExpr = setNames(data.frame(matrix(ncol = ncol(flowCore::exprs(tfm_fs[[1]]))+2, nrow = 0)),
                      c(colnames(tfm_fs), "samp", "tmr_pos"))

for(sample in names(cd8_files)){
  tmp.expr = as.data.frame(flowCore::exprs(tfm_fs[[sample]]))
  tmp.expr$samp = as.character(sample)
  tmp.expr$tmr_pos = c(rep("no", nrow(flowCore::exprs(cd8_files[[sample]]))),
                       if(!is.null(islet_tmr_files)) rep("islet", nrow(flowCore::exprs(islet_tmr_files[[sample]]))),
                       if(!is.null(insulin_tmr_files)) rep("insulin", nrow(flowCore::exprs(insulin_tmr_files[[sample]]))),
                       if(!is.null(viral_tmr_files)) rep("viral", nrow(flowCore::exprs(viral_tmr_files[[sample]]))))
  mergedExpr = rbind(mergedExpr, tmp.expr)
}

# Check for donors with too many/too few collected events
events_by_donor = mergedExpr %>%
  group_by(samp) %>%
  summarise(n_events = n())
View(events_by_donor)


```

## ============================
## Phenograph clustering + summarization
## ============================

```{r phenograph_clustering}
# Create vector of markers that will be used for all downstream clustering
clustering_markers <- toupper(key$desc)

# Remove GZMB and CD160 since they have high CVs in the reproducibility testing
clustering_markers = clustering_markers[!clustering_markers %in% c("GZMB","CD160")]

# Set up Phenograph function to use kd tree (from JinmiaoChenLab/Rphenograph github)
find_neighbors <- function(data, k){
  nearest <- RANN::nn2(data, data, k, treetype = "kd", searchtype = "standard")
  return(nearest[[1]])
}

Rpheno <- function(data, k=30){
  if(is.data.frame(data))
    data <- as.matrix(data)

  if(!is.matrix(data))
    stop("Wrong input data, should be a data frame or matrix!")

  if(k<1){
    stop("k must be a positive integer!")
  }else if (k > nrow(data)-2){
    stop("k must be smaller than the total number of points!")
  }

  message("Run Rphenograph starts:","\n",
          "  -Input data of ", nrow(data)," rows and ", ncol(data), " columns","\n",
          "  -k is set to ", k)

  cat("  Finding nearest neighbors...")
  t1 <- system.time(neighborMatrix <- find_neighbors(data, k=k+1)[,-1])
  cat("DONE ~",t1[3],"s\n", " Compute jaccard coefficient between nearest-neighbor sets...")
  t2 <- system.time(links <- Rphenograph:::jaccard_coeff(neighborMatrix))

  cat("DONE ~",t2[3],"s\n", " Build undirected graph from the weighted links...")
  links <- links[links[,1]>0, ]
  relations <- as.data.frame(links)
  colnames(relations)<- c("from","to","weight")
  t3 <- system.time(g <- igraph::graph.data.frame(relations, directed=FALSE))

  cat("DONE ~",t3[3],"s\n", " Run louvain clustering on the graph ...")
  t4 <- system.time(community <- igraph::cluster_louvain(g))
  cat("DONE ~",t4[3],"s\n")

  message("Run Rphenograph DONE, took a total of ", sum(c(t1[3],t2[3],t3[3],t4[3])), "s.")
  cat("  Return a community class\n  -Modularity value:", igraph::modularity(community),"\n")
  cat("  -Number of clusters:", length(unique(igraph::membership(community))))

  return(community)
}

# Run phenograph (using kd treetype) on each subject.
PhenographClust = function(fcs, clustering_markers) {
  exprs_mat = as.matrix(as.data.frame(flowCore::exprs(fcs))[,clustering_markers])
  RPvect = as.numeric(igraph::membership(Rpheno(data = exprs_mat)))
  return(RPvect)
}

mergedExpr$RPclust = unlist(lapply(all_tfm, PhenographClust, clustering_markers))

# Get summary of the number of clusters generated for each subject
n_pheno_clusts <- mergedExpr %>%
  group_by(samp) %>%
  summarize(k_clusters = max(RPclust))

View(n_pheno_clusts)

```

## ============================
## Summarize phenograph output
## ============================

```{r phenograph_cluster_mean_calculations}
# Calculate mean expression value of each marker for each phenograph cluster in each subject
RP_mean <- mergedExpr %>%
  dplyr::select(-tmr_pos) %>%
  group_by(samp, RPclust) %>%
  summarise_all(mean) %>%
  mutate(RPclust = as.character(RPclust))

# Calculate total CD8 mean expression for each subject
CD8_mean = mergedExpr %>%
  dplyr::select(-tmr_pos, -RPclust) %>%
  group_by(samp) %>%
  summarise_all(mean) %>%
  mutate(RPclust = "Total_CD8")

RP_mean = bind_rows(RP_mean, CD8_mean)

```


```{r phenograph_cluster_cell_counts}
# Count cells of each specificity in each phenograph cluster (from each sample)
RPtmr_counting = mergedExpr %>% 
  dplyr::select(samp, tmr_pos, RPclust) %>%
  group_by(samp, RPclust) %>%
  summarise(clust_size = n(), 
            None = sum(tmr_pos=="no"), 
            Islet = sum(tmr_pos=="islet"), 
            Insulin = sum(tmr_pos=="insulin"), 
            Virus = sum(tmr_pos=="viral"))

aggregate_counts = RPtmr_counting %>%
  dplyr::select(-RPclust) %>%
  group_by(samp) %>%
  summarise_all(sum) %>%
  rename_at(vars(-samp),function(name) paste0(name,"_tot"))

RPtmr_counting = RPtmr_counting %>%
  left_join(aggregate_counts) %>%
  mutate(pct_nonspec_in_clust = None/None_tot * 100,
         pct_islet_in_clust = Islet/Islet_tot * 100,
         pct_insulin_in_clust = Insulin/Insulin_tot * 100,
         pct_virus_in_clust = Virus/Virus_tot * 100)

# Save the clustering output
pheno_filename = paste0(fname_prefix_R, "_all_phenograph_data.RData")
save(mergedExpr, clustering_markers, RP_mean, RPtmr_counting, 
     file = pheno_filename)

```

## ============================
## tSNE clustering 
## ============================

You can skip this chunk (or try using UMAP) to speed things up.  It's unneccessary for DISCOV-R analysis but useful for visualization.

If you do skip this chunk, you'll need to pull the X_all_phenograph_data.RData file in at the beginning of the DISCOV-R script instead of the X_all_phenograph_tsne_data.RData file.

```{r tsne_clustering_by_subject}
# Create function to run tSNE
tsne_flowset <- function(fcs, selected_markers){
  tsne_data <- as.data.frame(flowCore::exprs(fcs))

  tsne_out <- tsne_data[,colnames(tsne_data) %in% selected_markers] %>%
    as.matrix %>%
    Rtsne(check_duplicates = F)

  tsne_data$tsne_1 = tsne_out$Y[,1]
  tsne_data$tsne_2 = tsne_out$Y[,2]

  return(tsne_data)
}

# Run tSNE on each sample
tsne_results <- lapply(all_tfm, tsne_flowset, selected_markers = clustering_markers)

# Tidy the results and merge them with the rest of the expression data.
tsne_results <- do.call("rbind", tsne_results)

if(setequal(tsne_results[,clustering_markers], mergedExpr[,clustering_markers])==T) {
  mergedExpr = mergedExpr %>%
    left_join(tsne_results)
  } else {
    print("The expression data from mergedExpr and tsne_results do NOT match!!")
  }

# Save the output with the other important exports.
tsne_filename = paste0(fname_prefix_R, "_all_phenograph_tsne_data.RData")
save(mergedExpr, clustering_markers, RP_mean, RPtmr_counting,
     file = tsne_filename)

#load(tsne_filename)
```


```{r plot_tsne_data}
# Plot insulin Tmr+ cells on each subject's tSNE plot
InsBySubj = lapply(unique(mergedExpr$samp), function(x) {
  ggplot(data = mergedExpr[mergedExpr$samp == x,], aes(x=tsne_1, y = tsne_2, color = factor(RPclust)))+
    geom_point(size=1, alpha = 0.4)+
    geom_point(data = mergedExpr[mergedExpr$samp == x & mergedExpr$tmr_pos == "insulin",],
               size=2, color = "black", alpha = .7, shape = 15)+
    scale_color_manual(values = colorRampPalette(cb_pal)(length(unique(mergedExpr$RPclust[mergedExpr$samp == x]))))+
    labs(title=x, x="tsne 1", y="tsne 2")+
    guides(color=guide_legend(ncol=2))+
    theme(text = element_text(size=14))
})
grid_Ins = arrangeGrob(grobs = InsBySubj, ncol = 5)
insulin_filename = paste0(fname_prefix_tsne, "_Insulin_tSNE_byDonor.pdf")
ggsave(insulin_filename, grid_Ins, width = 29, height = 49)

# Plot islet Tmr+ cells on each subject's tSNE plot
IslBySubj = lapply(unique(mergedExpr$samp), function(x) {
  ggplot(data = mergedExpr[mergedExpr$samp == x,], aes(x=tsne_1, y = tsne_2, color = factor(RPclust)))+
    geom_point(size=1, alpha = 0.4)+
    geom_point(data = mergedExpr[mergedExpr$samp == x & mergedExpr$tmr_pos == "islet",],
               size=2, color = "black", alpha = 0.7, shape = 17)+
    scale_color_manual(values = colorRampPalette(cb_pal)(length(unique(mergedExpr$RPclust[mergedExpr$samp == x]))))+
    labs(title=x, x="tsne 1", y="tsne 2")+
    guides(color=guide_legend(ncol=2))+
    theme(text = element_text(size=14))
})
grid_Isl = arrangeGrob(grobs = IslBySubj, ncol = 5)
iset_filename = paste0(fname_prefix_tsne, "_Islet_tSNE_byDonor.pdf")
ggsave(iset_filename, grid_Isl, width = 29, height = 49)

# Plot viral Tmr+ cells on each subject's tSNE plot
VirBySubj = lapply(unique(mergedExpr$samp), function(x) {
  ggplot(data = mergedExpr[mergedExpr$samp == x,], aes(x=tsne_1, y = tsne_2, color = factor(RPclust)))+
    geom_point(size=1, alpha = 0.4)+
    geom_point(data = mergedExpr[mergedExpr$samp == x & mergedExpr$tmr_pos == "viral",],
               size=2, color = "black", alpha = 0.7, shape = 8)+
    scale_color_manual(values = colorRampPalette(cb_pal)(length(unique(mergedExpr$RPclust[mergedExpr$samp == x]))))+
    labs(title=x, x="tsne 1", y="tsne 2")+
    guides(color=guide_legend(ncol=2))+
    theme(text = element_text(size=14))
})
grid_Vir = arrangeGrob(grobs = VirBySubj, ncol = 5)
viral_filename = paste0(fname_prefix_tsne, "_Viral_tSNE_byDonor.pdf")
ggsave(viral_filename, grid_Vir, width = 29, height = 49)

# Save plot objects so you can go back and print out a single subject's tSNE 
plot_filename = paste0(fname_prefix_R, "_tsne_plot_objects.Rdata")
save(InsBySubj, IslBySubj, VirBySubj, file = plot_filename)

```

